Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-10-31 paper/paper.tex
semantics progress

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}

@@

\newcommand{\bang}{~\textbf{!}~}
% Tentatively using uparrow for linear to non-linear to be in line with
% Pfennings adjoint functional programming
\newcommand{\ltonl}[1]{~\uparrow #1}
\newcommand{\nil}{\mathsf{nil}}
\newcommand{\cons}{\mathsf{cons}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\where}{\mathsf{where}}
\newcommand{\Trace}{\mathsf{Trace}}
\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
\newcommand{\internalize}[1]{\lceil#1\rceil}

@@

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

@@

@@

% TODO this should be better formatted but can't use listings with \otimes ^{\ast}
\begin{figure}
\begin{align*}
\data &~A^{*} : L~\where\\
      & \nil : \ltonl {A^{*}} \\
      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

The arrow, $\uparrow$, wrapping these constructors denotes
an interaction between the linear and nonlinear parts of the theory.
In short, a nonlinear term
$M : \ntonl A^{*}$ denotes a linear term in the empty context
$\cdot \vdash M : A^{*}$. Later in
this paper, we will further discuss the behavior of $\uparrow$.
At this moment,
it suffices to read these arrows as allow us to use the reuse the names
``$\nil$'' and ``$\cons$''
as symbols for constructors. That is, the $\nil$ constructor is not linearly
consumed at usage time

Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
star constructors. In this proof tree we expand some syntactic details that are often
left implicit. For instance, every linear context is parameterized by a nonlinear
context, which is syntactically separate by a semicolon. If the nonlinear
context is empty, we may often omit it; however, in this figure we make this
distinction explicit. Further consider the leaves of the tree. First, the
leaves involving the constructors are nonlinear terms, and thus they occur in
the nonlinear empty context. Then we explicit cast from
an nonlinear $uparrow$-wrapped type into a linear one. Such casts may often be
made implicitly, such as in $\cref{fig:kleeneabstractproof}$.

@@

Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
$(a^{*} \otimes b) \oplus c$, we show in \cref{fig:nfaderivation} that \stingquote{ab} is
accepted by the NFA representation of this regular expression. This is denoted by
showing that \stringquote{ab} matches the trace grammar when the index is the
initial state of the NFA.\ Further, as we
will see in LATER SECTION, we can derive an isomorphism between parses of the
regular expression and accepting traces of their corresponding NFA.

@@

@@

\begin{figure}
  \[
  \begin{array}{rrcl}
    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
                                                  \mathsf{Int} \pipe
                                                  \mathsf{Empty} \pipe
                                                  \mathsf{Unit} \pipe
                                                  \mathsf{SPF}\,X\pipe U \pipe L
                                                  \pipe \ltonl A \pipe \sum_{x:X} Y
                                                  \pipe \prod_{x:X} Y\pipe M =_A
                                                  N \\
    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
    \textrm{linear types} & A , B , C & ::= &
    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M \pipe
    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
  \end{array}
  \]
  \caption{Overview of Types}
\end{figure}

@@

\subsection{Subtheories}
Each choice of connective has direct implications on the expressivity of
\theoryabbv. In particular, the set of connectives used will determine where the
grammar semantics is placed in
the Chomsky hierarchy.

\paragraph{Regular Expressions}
We may realize regular expression as a subtheory
of our language by restricting the type constructors to the
linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
languages recognized by these are exactly the regular languages --- the lowest
on the Chomsky hierarchy.

\paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
left-fixed point rather than only Kleene star.
Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
expressions, which are known to be equivalent to context-free grammars
\cite{leiss,krishnaswami_typed_2019}.

\paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
regular grammars and context-free grammars, respectively; however, by including
the LNL dependent types we may actually express the entirety of the Chomsky
hierarchy. Through use of the LNL dependent types, we may encode indexed
grammars, which are known to be properly between context-free and
context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
calculus, which of course induces a semantics of unrestricted grammars.

\paragraph{Indexed Grammars}
The LNL dependent types can be used to encode Aho's indexed grammars
\cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
things with this non-linear index, as we have unrestricted access to the index
whereas Aho's grammars can only manipulate it in a particular way.

\steven{Elaborate on these, and note that they are bw context free and context
  sensitive on the Chomsky hierarchy}

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}

\steven{this section is mostly word vomit, need to edit}

\subsection{Grammar Properties}
\steven{Mention that $\top$ and $\String$ are iso. Highlight this as a needed axiom}

%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
\paragraph{Unambiguity}
In the grammars model, we say that a grammar is unambiguous if there is at most
one parse tree for any input string. There are several ways to capture this
concept inside of the theory of \theoryabbv.

In the presence of distributive coproducts (which we probably add a) the following are equivalent:
\begin{itemize}
  \item $A$ is an unambiguous grammar
  \item The morphism $! : A \to \top$ is a monomorphism
  \item Given a grammar $B$, all morphisms $B \to A$ are equal
  \item $A \& A \cong A$
\end{itemize}

\paragraph{Parseability}
We say $A$ is \emph{parseable} if there exists a type $B$ and a map
$\top \to A \oplus B$. Because $\top \cong \String$, this says that any string
can be parsed into either an $A$ or a $B$.

If $\top \cong A \oplus B$, call $A$ \emph{unambiguously parseable}.

\paragraph{Decidability}
In the case that $A$ is parseable with respect to $\neg A$ --- i.e.
there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.

\paragraph{Equivalence}
Grammars $A$ and $B$ are \emph{weakly equivalent} if there exists maps
$f : A \to B$ and $g : B \to A$.

If $f$ and $g$ form an isomorphism, then $A$ and $B$ are \emph{strongly
  equivalent}.
\steven{TODO cite that these are the same as Chomsky's notions of equivalence}

The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
a bijective translation of parses. In other words, two grammars are isomorphic
if they capture the same structural descriptions of strings, which is
precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}. In works like
\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of equivalence arise
naturally.

@@

\paragraph{Regex-NFA/Thompson}

\steven{TODO put regex-NFA iso here. Call it Thompson's construction?. Mention
  its mechanized}

\paragraph{Determinization of an NFA}
\newcommand{\Agda}{Formalized in Agda}
\begin{theorem}[Determinization, \Agda]
  Given an NFA $N$, there exists a DFA $D$ such that
  $\Trace_{N}~N.\mathsf{init}$ is weakly equivalent to $\Trace_{D}~D.\mathsf{init}$.
\end{theorem}

\begin{proof}
  We present a variant of the classic powerset construction CITE. First, we build
  the DFA $D$. Define the
  states of $D$ to be the $\mathbb{P}_{\epsilon}(N.\mathsf{states})$ --- the type of $\epsilon$-closed subsets of
  $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
  from $N$. Construct the initial state of $D$ as the
  $\epsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
  of $D$ to send the subset $X$ under the character $c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $c$.

@@

If the NFA trace is a labeled transition $s \overset{c}{\to} s'$ such that
$s \in X$, then we first decompose the trace to a parse of the character $c$ and a
trace starting from $s'$ --- that is, to a term of type
$c \otimes \Trace_{N}(s')$. We then inductively call $f$ on traces out of $s'$
to get traces in a parse of $c$ concatenated with a family of traces
parametrized by $\epslion$-closed subsets containing $s'$.
. We instantiate this family at $\delta_{c}(X)$, as
$s'$ is reachable from $s$ by a single transition labelled by $c$, so
$s' \in \overline{\epsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
to get a trace in $D$ rooted at $X$.

@@

Let's turn our heads for a moment to the content behind the traces in $D$.
Traces from subset $X$ to subset $Y$ denote the \emph{existence} of some trace
in the $N$ from some $x \in X$ to some $y \in Y$. In general it is not clear how
to extract out a choice of trace in $N$, which is what $g$ is asking of us.
However, we may further take the states and transitions of $N$ to be ordered.
This allows us to
induce an order on traces in $D$ from $X$ to $Y$. From these orderings we can then
define the necessary choice functions on each type by taking the smallest with
respect to the ordering.

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
The Dyck grammar $D$ for balanced parens (summarize what it is + citation)

>>>>

\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\mathsf{string}}
\newcommand{\charg}{\mathsf{char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$}

\newcommand{\isTy}{\textrm{ type}}
\newcommand{\isCtx}{\textrm{ ctx}}
\newcommand{\isSmall}{\textrm{ small}}
\newcommand{\isLinTy}{\textrm{ lin. type}}
\newcommand{\isLinCtx}{\textrm{ lin. ctx.}}

@@

\newcommand{\bang}{~\textbf{!}~}
% Tentatively using uparrow for linear to non-linear to be in line with
% Pfennings adjoint functional programming
\newcommand{\ltonl}[1]{~\uparrow #1}
\newcommand{\nil}{\mathsf{nil}}
\newcommand{\cons}{\mathsf{cons}}
\newcommand{\epscons}{\epsilon\mathsf{cons}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\where}{\mathsf{where}}
\newcommand{\Trace}{\mathsf{Trace}}
\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}
\newcommand{\internalize}[1]{\lceil#1\rceil}

@@

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\nathan}[1]{\ifdraft{\color{green}[{\bf Nathan says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

@@

\author{Nathan Varner}
\affiliation{\department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{nvarner@umich.edu}

@@

\begin{figure}
\begin{align*}
\data &~A^{*} : L~\where\\
      & \nil : \ltonl {A^{*}} \\
      & \cons : \ltonl {(A \lto A^{*} \lto A^{*})}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

The arrow, $\uparrow$, wrapping these constructors denotes
an interaction between the linear and nonlinear parts of the theory.
In short, a nonlinear term
$M : \ltonl A^{*}$ denotes a linear term in the empty context
$\cdot \vdash M : A^{*}$. Later in
this paper, we will further discuss the behavior of $\uparrow$.
At this moment,
it suffices to read these arrows as allow us to use the reuse the names
``$\nil$'' and ``$\cons$''
as symbols for constructors. That is, the $\nil$ constructor is not linearly
consumed at usage time

Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
star constructors. In this proof tree we expand some syntactic details that are often
left implicit. For instance, every linear context is parameterized by a nonlinear
context, which is syntactically separated by a semicolon. If the nonlinear
context is empty, we may often omit it; however, in this figure we make this
distinction explicit. Further consider the leaves of the tree. First, the
leaves involving the constructors are nonlinear terms, and thus they occur in
the nonlinear empty context. Then we explicit cast from
an nonlinear $\uparrow$-wrapped type into a linear one. Such casts may often be
made implicitly, such as in $\cref{fig:kleeneabstractproof}$.

@@

Just as \cref{fig:kleenestarderivation} shows the \stringquote{ab} matches
$(a^{*} \otimes b) \oplus c$, we show in \cref{fig:nfaderivation} that \stringquote{ab} is
accepted by the NFA representation of this regular expression. This is denoted by
showing that \stringquote{ab} matches the trace grammar when the index is the
initial state of the NFA.\ Further, as we
will see in LATER SECTION, we can derive an isomorphism between parses of the
regular expression and accepting traces of their corresponding NFA.

@@

\steven{TODO the example figures need to changed to not use $x$'s for linear variables}

@@

\begin{figure}
  \[
  \begin{array}{rrcl}
    \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
    \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
                                                  \mathsf{Int} \pipe
                                                  \mathsf{Empty} \pipe
                                                  \mathsf{Unit} \pipe
                                                  \mathsf{SPF}\,X\pipe U \pipe L
                                                  \pipe \ltonl A \pipe \sum\limits_{x:X} Y
                                                  \pipe \prod\limits_{x:X} Y\pipe M =_A
                                                  N \\
    \textrm{linear contexts} & \Delta & ::= & \cdot \pipe \Delta , a:A\\
    \textrm{linear types} & A , B , C & ::= &
    c \pipe I \pipe A \otimes B \pipe \bigoplus\limits_{x:X} A \pipe \mu A_f M \pipe
    A \lto B \pipe B \tol A \pipe \bigwith\limits_{x:X} A \pipe \equalizer {a}{f}{g}\\
  \end{array}
  \]
  \caption{Overview of Types}
\end{figure}

@@

% \paragraph{Regular Expressions}
% We may realize regular expression as a subtheory
% of our language by restricting the type constructors to the
% linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
% languages recognized by these are exactly the regular languages --- the lowest
% on the Chomsky hierarchy.

% \paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
% to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
% left-fixed point rather than only Kleene star.
% Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
% expressions, which are known to be equivalent to context-free grammars
% \cite{leiss,krishnaswami_typed_2019}.

% \paragraph{Beyond Context-Free Grammars} The previous two subtheories induce as semantics
% regular grammars and context-free grammars, respectively; however, by including
% the LNL dependent types we may actually express the entirety of the Chomsky
% hierarchy. Through use of the LNL dependent types, we may encode indexed
% grammars, which are known to be properly between context-free and
% context-sensitive grammars within the Chomsky hierarchy \cite{AhoIndexed}. We
% will further see in \cref{subsubsec:tm} how to use dependence to encode Turing machines internal to our
% calculus, which of course induces a semantics of unrestricted grammars.

% \paragraph{Indexed Grammars}
% The LNL dependent types can be used to encode Aho's indexed grammars
% \cite{AhoIndexed} internal to \theoryabbv. Although, we can define strictly more
% things with this non-linear index, as we have unrestricted access to the index
% whereas Aho's grammars can only manipulate it in a particular way.

% \steven{Elaborate on these, and note that they are bw context free and context
%   sensitive on the Chomsky hierarchy}

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to conducting formal
gramamr theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname. By encoding well-established formal grammar concepts, we ensure
that our framework remains grounded in the
foundational principles of formal language theory while opening the door to
compositional formal verification of parsers.

\subsection{Grammar Properties}

\steven{Need an intro for this subsection}

%% 1. Basic concepts: unambiguity, parseability, weak and strong equivalence
\paragraph{Unambiguity}
Ordinarily, we think of a formal grammar as unambiguous when there is at most
one parse tree for any input string; however, this description of unambiguity is
a semantic property that is not expressible in the syntax of \theoryabbv. Rather
than capturing this uniqueness aspect directly, we provide several notions
internal to \theoryabbv that imply unambiguity in the grammars model.

We say that $A$ is \emph{unambiguous} if the unique map $! : A \to \top$ to the
terminal object is a monomorphism. Further, in the presence of distributive coproducts, the following are equivalent:
\begin{itemize}
  \item $A$ is an unambiguous grammar
  \item Given a grammar $B$, all morphisms $B \to A$ are equal
  \item $A \& A \cong A$
\end{itemize}

\paragraph{A Grammar of Strings}
Define $\charg$ to be the disjunction of characters in the alphabet $\Sigma$,
$\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings,
$\stringg = \charg^{*}$. In the grammars model, we may construct an
isomorphism between $\top$ and $\stringg$. Thus when building parsers, we may wish to enforce this
syntactically by including $\top \cong \stringg$ as an axiom.

\paragraph{Parseability}
We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
$\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
refutation.

\paragraph{Decidability}
In the case that $A$ is parseable with respect to $\neg A$ --- i.e.\
there's a map $\top \to A \oplus \neg A$ --- then call $A$ \emph{decidable}.

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} by
Chomsky if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

Internal to \theoryabbv, we define grammars $A$ and $B$ to be \emph{weakly equivalent} if there exists maps
$f : A \to B$ and $g : B \to A$. Note that in the grammars model, this implies the expected equality on
the set of accepted strings.

If $f$ and $g$ form an isomorphism, then $A$ and $B$ defined to be \emph{strongly
  equivalent}. This implies Chomsky's notion of strong equivalence, as for any
input string $w$
$A \cong B$ induces a bijective translation of parses $A w \cong B w$.
When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of strong equivalence
emerge naturally.

@@

Regular expressions
are among one of the most studied and
well-understood concepts in all of computer science. Pratically, they are
essential tools for lexical analysis and many text
processing tasks. So they are a natural concept to encoding into our
domain-specific grammars language.

Nearly every theoretical discussion of regular expressions is accompanied by a
discussion of finite automata, as it is well known that regular expressions and finite automata are
equivalent as grammar formalisms. In this section, we make the equivalences
between regular expressions and finite automata precise by encoding two theorems
from classical language theory: (1) every regular expression is strongly equivalent
to the traces through some nondeterministic finite automaton (NFA); and, (2) for
every NFA, there is some deterministic finite automaton (DFA) such that the type
of traces through the NFA is weakly equivalent to the type of traces through the DFA.

\newcommand{\states}{\mathsf{states}}
\newcommand{\labelt}{\mathsf{label}}
\newcommand{\transitions}{\mathsf{transitions}}
\newcommand{\epstransitions}{\epsilon\mathsf{transitions}}
\newcommand{\isAcc}{\mathsf{isAcc}}
\newcommand{\init}{\mathsf{init}}
\newcommand{\src}{\mathsf{src}}
\newcommand{\dst}{\mathsf{dst}}
\newcommand{\epssrc}{\epsilon\mathsf{src}}
\newcommand{\epsdst}{\epsilon\mathsf{dst}}

\paragraph{Regular Expressions} The theory of regular expressions emerges as the
subtheory of \theoryabbv generated by
$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.

\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw the demonstration of
one particular NFA corresponding to the regular expression
$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
through this
NFA, as
presented in \cref{fig:nfainductive}, inspects the concrete structure of the
NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
luxury of inspecting a concrete defintion. Therefore, we abstract over the
specification of an arbitrary NFA and define traces with respect to that
specification.

Let the data of an NFA $N$ be given by the following nonlinear data,
\begin{itemize}
\item $N.\states$ --- a finite set of states
\item $N.\init$ --- the initial state
\item $N.\isAcc$ --- a decidable proposition over $N.\states$ denoting which
        states are accepting
\item $N.\transitions$ --- a finite set of labeled transitions
\item $N.\labelt$ --- a mapping from $N.\transitions$ to the character labeling
        the transition
\item $N.\src$ --- a mapping from $N.\transitions$ to source state of a transition
\item $N.\dst$ --- a mapping from $N.\transitions$ to destination state of a transition
\item $N.\epstransitions$ --- a finite set of $\epsilon$-transitions
\item $N.\epssrc$ --- a mapping from $N.\epstransitions$ to source state of a $\epsilon$-transition
\item $N.\epsdst$ --- a mapping from $N.\epstransitions$ to destination state of a $\epsilon$-transition
\end{itemize}

For the above definition of an NFA $N$, in \cref{fig:nfatrace} we define a linear type of traces
through $N$. Traces in $N$ may be built through one of three constructors. We
may terminate a trace at an accepting state with the constructor $\nil$. If we
had a trace beginning at the destination state of a transition, then we may use
the $\cons$ constructor to linearly combine that trace with a parse of the label
of the transition to build a trace beginning at the source of the transition.
Finally, if we had a trace beginning at the destination of an
$\epsilon$-transition then we may use $\epscons$ to pull it back along the
$\epsilon$-transition and construct a trace beginning at the source of the $\epsilon$-transition.

\begin{figure}
\begin{align*}
\data &~\Trace_{N} : (s : N.\states) \to L~\where\\
      & \nil : \ltonl {\left( \LinPiTy {s} {N.\states} {\LinPiTy {\mathsf{acc}} {N.\isAcc(s)} {\Trace_{N}~s}} \right)}  \\
      & \cons : \ltonl {\left( \LinPiTy {t} {N.\transitions} {(N.\labelt~t) \lto \Trace_{N}~(N.\dst(t)) \lto \Trace_{N}(N.\src(t))} \right)} \\
      & \epscons : \ltonl {\left( \LinPiTy {t} {N.\epstransitions} {\Trace_{N}~(N.\epsdst(t)) \lto \Trace_{N}(N.\epssrc(t))} \right)}
\end{align*}
\caption{Traces through the NFA $N$ as an indexed inductive type}
\label{fig:nfatrace}
\end{figure}

\paragraph{Determinstic Finite Automata}
Just as above, we provide a specification for an arbitrary DFA.\ A DFA $D$ is
given by the following data,

\begin{itemize}
\item $D.\states$ --- a finite set of states
\item $D.\init$ --- the initial state
\item $D.\isAcc$ --- a decidable proposition over $D.\states$ denoting which
      states are accepting
\item $D.\delta$ --- a transition function. Given a character $c$ and a state
        $s$, $D.\delta_{c}(s)$ is the state for which there is a transition $s \overset{c}{\to} D.\delta_{c}(s)$
\end{itemize}

The linear type of traces through $D$ is given in \cref{fig:dfatrace}. These
traces are either terminated in an accepting state with the $\nil$ constructor,
or the $\cons$ constructor builds a trace out of state $s$ by linearly combining
a parse of some character $c$ with a
trace out of the state $D.\delta_{c}(s)$.

\begin{figure}
\begin{align*}
\data &~\Trace_{D} : (s : D.\states) \to L~\where\\
      & \nil : \ltonl {\left( \LinPiTy {s} {D.\states} {\LinPiTy {\mathsf{acc}} {D.\isAcc(s)} {\Trace_{D}~s}} \right)}  \\
      & \cons : \ltonl {\left( \LinPiTy {c} {\Sigma} {\LinPiTy {s} {D.\states} {c \lto \Trace_{D}~(D.\delta_{c}(s)) \lto \Trace_{D}(s)}} \right)}
\end{align*}
\caption{Traces through the DFA $D$ as an indexed inductive type}
\label{fig:dfatrace}
\end{figure}

\newcommand{\Agda}{Formalized in Agda}
\begin{theorem}[Determinization, \Agda]
  Given an NFA $N$, there exists a DFA $D$ such that
  $\Trace_{N}~N.\mathsf{init}$ is weakly equivalent to $\Trace_{D}~D.\mathsf{init}$.
\end{theorem}

\begin{proof}
  We present a variant of the Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
  the DFA $D$. Define the
  states of $D$ to be the $\mathbb{P}_{\epsilon}(N.\mathsf{states})$ --- the type of $\epsilon$-closed subsets of
  $N.\mathsf{states}$. A subset is accepting in $D$ if it contains an accepting state
  from $N$. Construct the initial state of $D$ as the
  $\epsilon$-closure of $N.\mathsf{init}$. Lastly define the transtion function
  of $D$ to send the subset $X$ under the character $c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $c$.

@@

If the NFA trace is a labeled transition $s \overset{c}{\to} s'$ such that
$s \in X$, then we first decompose the trace to a parse of the character $c$ and a
trace starting from $s'$ --- that is, to a term of type
$c \otimes \Trace_{N}(s')$. We then inductively call $f$ on traces out of $s'$
to get traces in a parse of $c$ concatenated with a family of traces
parametrized by $\epsilon$-closed subsets containing $s'$.
. We instantiate this family at $\delta_{c}(X)$, as
$s'$ is reachable from $s$ by a single transition labelled by $c$, so
$s' \in \overline{\epsilon}\overline{c}{X} = D.\delta_{c}(X)$. Therefore we can
linearly combine the parse of character $c$ and the trace out of $D.\delta_{c}(X)$
to get a trace in $D$ rooted at $X$.

@@

For a moment, let us turn our attention to the content behind the traces in $D$.
Traces from subset $X$ to subset $Y$ denote the \emph{existence} of some trace
in the $N$ from some $x \in X$ to some $y \in Y$. In general it is not clear how
to extract out a choice of trace in $N$, which is what $g$ is asking of us.
However, we may further take the states and transitions of $N$ to be ordered.
This allows us to
induce an order on traces in $D$ from $X$ to $Y$. From these orderings we can then
define the necessary choice functions on each type by taking the smallest with
respect to the ordering.

@@

\begin{theorem}[Thompson's Construction, \Agda]
  Given a regular expression $R$, there exists an NFA $N$ such that $R$ is
  strongly equivalent to $\Trace_{N}(N.\init)$.
\end{theorem}

\begin{proof}
In this proof, we present a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968} in which we induct on the
structure of $R$ to construct strongly equivalent NFAs. We will sketch this
proof by walking through some of the inductive cases, the remaining cases follow
similar reasoning and the details may be found in the formalized artifact.

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial] (0) {$0$};
    \node[state, right of=0, accepting] (1) {$1$};

    \path[->] (0) edge[above] node{$c$} (1);
  \end{tikzpicture}
  \caption{NFA $N_{c}$ for a single character $c$}
  \label{fig:litnfa}
\end{figure}

First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
and show they are mutually inverse.

Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,

\[
  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
\]

Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
to the linear unit $I$. Then we must handle traces beginning in state 0, which
are necessarily built via $\cons$. We decompose such a trace into a parse of the
character $c$ and a trace beginning in state 1. We inductively transform the
trace beginning in state 1 into a term of type $I$. Finally, we use the right
unitor to turn a term of type $c \otimes I$ into one of type $c$.

\[\begin{tikzcd}
    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
    {c \otimes I} &
    c
	\arrow["{id\otimes g}", from=1-1, to=1-2]
	\arrow["{}", from=1-2, to=1-3]
\end{tikzcd}\]

Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
identity on $c$. Additionally, we can show that $f$ carries the structure of an
indexed algebra homomorphism for the family of functors defining the trace type
$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
homomorphism. $f \circ g$ and the identity map are both then algebra
homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
initial algebra for the relevant family of functors, there is only one
homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
$f$ and $g$ are mutually inverse and the regular expression is strongly
equivalent to the traces through the NFA in this case.
